Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-03-01 latex/MPP21/paper.tex
added abstract and dummy plots for matmul performance

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{tikz,pgfplots}
\usetikzlibrary{positioning}

@@

\newcommand{\datafile}{}

@@

\begin{abstract}
\GB{TBD}
\end{abstract}

@@

Matrix multiplication in particular is a bottleneck computation for many neural networks.
Forward and backward propagation in training the weights of fully connected layers is a matrix multiplication with dimensions given by the sizes of the layers and the size of the batch.
Training convolutional and other types of layers can also be cast as matrix multiplication, either via monolithic multiplications or batches of smaller multiplications \cite{CW+14,GB+19}.
In this paper we focus on Multi-Layer Perceptron (MLP) networks that rely on a sequence of fully connected layers \cite{HSW89}.
Because the sizes of the layers in MLP networks continue to grow, various techniques have been used to reduce the computational demands of the training phase.
For example, low-rank tensor approximation of the weights can reduce both memory and computation \cite{NPOV15}, and fast matrix multiplication (the Strassen-Winograd algorithm) has been applied to the bottleneck matrix multiplications \cite{KAA20}.

Our contribution is the use of APA matrix multiplication algorithms to address this problem and accelerate training.
In particular we
\begin{enumerate}
	\item curate a collection of well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them;
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to \GB{$Y\times$}.
\end{enumerate}

@@

\GB{should we add an accuracy plot in this section?}

@@

\Cref{tab:algs} shows the key performance and accuracy properties of the APA algorithms we consider.
Each row corresponds to an algorithm, and the first row includes the classical algorithm for comparison.
For all algorithms, we assume only 1 recursive step is used, though the speedup and error for more steps can be readily calculated from the parameters.
The first column gives the reference where the algorithm was first specified.
The second block column demonstrates the possible performance improvement, where speedup is calculated as $(mnk/r - 1)\cdot100\%$ for $\dims{m,n,k}$ and rank $r$.
The third block column shows error parameters, with error calculated as $2^{-d\cdot\sigma/(\sigma+\varphi)}$ with $d=23$, corresponding to single precision.

@@

@@

>>>>

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{ifthen}

% cleveref
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\crefname{section}{\S}{Sections}
\crefname{figure}{Fig.}{Figs.}
\crefname{table}{Table}{Tables}

% tikz
\usepackage{tikz,pgfplots}
\usetikzlibrary{positioning}

@@

% macros for plots
\newcommand{\datafile}{}
\newcommand{\threads}{}

@@

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for shared-memory platforms to obtain up to \GB{Z\%} speedups over the fastest implementation of the classical algorithm.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no signification reduction in the training or testing error.
Our performance results on a large MLP network show overall performance improvements of up to \GB{W\%}.
\end{abstract}

@@

Matrix multiplication in particular is a bottleneck computation for many neural networks.
Forward and backward propagation in training the weights of fully connected layers requires matrix multiplication with dimensions given by the sizes of the layers and number of batch samples.
Training convolutional and other types of layers can also be cast as matrix multiplication, either via monolithic multiplications or batches of smaller multiplications \cite{CW+14,GB+19}.
In this paper we focus on Multi-Layer Perceptron (MLP) networks that rely on a sequence of fully connected layers \cite{HSW89}.
Because the sizes of the layers in MLP networks continue to grow, various techniques have been used to reduce the computational demands of the training phase.
For example, low-rank tensor approximation of the weights can reduce both memory and computation \cite{NPOV15}, and fast matrix multiplication (the Strassen-Winograd algorithm) has been applied to the bottleneck matrix multiplications \cite{KAA20}.

Our contribution is the use of APA matrix multiplication algorithms to address this problem and accelerate training.
In particular we
\begin{enumerate}
	\item curate a collection of well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them;
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to \GB{$Y\times$} over the use of classical matrix multiplication.
\end{enumerate}

@@

@@

\Cref{tab:algs} shows the key performance and accuracy properties of the APA algorithms we consider.
Each row corresponds to an algorithm, and the first row includes the classical algorithm for comparison.
For all algorithms, we assume only 1 recursive step is used, though the speedup and error for more steps can be readily calculated from the parameters.
The first column gives the reference where the algorithm was first specified.
The second block column demonstrates the possible performance improvement, where speedup is calculated as $(mnk/r - 1)\cdot100\%$ for $\dims{m,n,k}$ and rank $r$.
The third block column shows error parameters, and the error is calculated as $2^{-d\cdot\sigma/(\sigma+\varphi)}$ with $d=23$, corresponding to single precision.

@@

\begin{figure}
\centering
\renewcommand{\datafile}{data/matmul_seq.dat}
\renewcommand{\threads}{1}
\input{plots/matmul_perf}
\caption{Single-threaded matrix multiplication performance}
\label{fig:matmul_seq}
\end{figure}

@@

\begin{figure}
\centering
\renewcommand{\datafile}{data/matmul_par.dat}
\renewcommand{\threads}{12}
\input{plots/matmul_perf}
\caption{Multithreaded matrix multiplication performance}
\label{fig:matmul_seq}
\end{figure}
